\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
adjust cif language

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\

@@

\caption{Comparison of various R packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

  1. case-base sampling; 
  2. estimation of the smooth hazard function; 
  3. estimation of the CIF.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from \pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data. We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

plot(survfit(cox_model, newdata = new_data),
     ylab = "Cumulative incidence (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
```

Moreover, we can also compute confidence intervals for the CIF using the method \code{confint.absRiskCB}.

@@

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the CIF estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

@@

par(mfrow = c(1, 2))
# 1. Control
tt_sub <- subset(tt, cov_prof == "Control group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        ylab = "Cumulative incidence (%)",
        xlab = "Years since randomization",
        main = "Control group"
    )
)

@@

# 2. Screening
tt_sub <- subset(tt, cov_prof == "Screening group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        # ylab = "Cumulative incidence (%)",
        ylab = "",
        xlab = "Years since randomization",
        main = "Screening group"
    )
)

@@

We then compare the cumulative incidence estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

@@

In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

@@

As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the CIF. Accordingly, our package provides functions for each step. Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface. This interface should look familiar to new users. Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.

\begin{widetable}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{widetable}

@@

In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners. The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}. Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the survival function that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the survival function that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Risk}  \\
\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Estimates}           \\

@@

\caption{Comparison of various R packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Risk estimates}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

  1. case-base sampling; 
  2. estimation of the smooth hazard function; 
  3. estimation of the survival function.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
\begin{equation}\label{eqn:CDF}
F(t) = 1 - S(t).
\end{equation}

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

plot(survfit(cox_model, newdata = new_data),
     ylab = "Risk probability (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
```

Moreover, we can also compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.

@@

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

@@

par(mfrow = c(1, 2))
# 1. Control
tt_sub <- subset(tt, cov_prof == "Control group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        ylab = "Risk probability (%)",
        xlab = "Years since randomization",
        main = "Control group"
    )
)

@@

# 2. Screening
tt_sub <- subset(tt, cov_prof == "Screening group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        # ylab = "Risk probability (%)",
        ylab = "",
        xlab = "Years since randomization",
        main = "Screening group"
    )
)

@@

We then compare the risk estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

@@

In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating risk functions using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

@@

As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function. Accordingly, our package provides functions for each step. Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface. This interface should look familiar to new users. Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.

\begin{widetable}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Survival curve & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{widetable}

@@

In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth survival functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners. The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}. Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.
